The Inconsistency in G\"odel's Ontological Argument: A Success Story for Automated Theorem Proving in Metaphysics

### 2016-06-24 Papers/2016/KI/main.tex
update

\title{The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics}
\author{Christoph Benzm\"uller\inst{1}\thanks{This work has been supported by
    the German Research Foundation DFG under grants BE2501/9-1,2} \and Bruno Woltzenlogel Paleo\inst{2} 
}

@@

\begin{abstract}
  This paper discusses the discovery of the inconsistency in G\"odel's
  ontological argument as a success story for artificial
  intelligence. Despite the popularity of the argument since the
  appearance of G\"odel's manuscript in the early 1970's, the
  inconsistency of the axioms used in the argument remained unnoticed
  until 2013, when it was detected automatically by the higher-order
  theorem prover \textsc{Leo-II}.  Understanding and verifying the
  refutation generated by the prover turned out to be a time-consuming
  task. Its completion, as reported here, required the
  reconstruction of the refutation in the Isabelle proof assistant.% , and it
  % also led to a novel and more efficient way of automating
  % higher-order modal logic \textbf{S5} with a universal accessibility
  % relation. Furthermore, the development of an improved syntactical
  % hiding for the utilized logic embedding technique allows the refutation to be
  % presented in a human-friendly way, suitable for non-experts in the
  % technicalities of higher-order theorem proving. This brings us a
  % step closer to wider adoption of logic-based artificial intelligence
  % tools by philosophers.
  This paper is short version our contribution to IJCAI 2016.
\end{abstract}

@@

The main novel contribution reported here is a detailed
analysis (in various modal logics) of the inconsistency of G\"{o}del's
original version of the axioms used in his manuscript
\cite{GoedelNotes}. The extraction, reconstruction and
verification of an informal, human intuitive argument has been an open
problem since the first detection of this inconsistency by
Benzm\"uller and Woltzenlogel-Paleo \cite{C40} with the
\textsc{Leo-II} prover.  The verified refutation (discussed in
\S\ref{sec:inconsistency}) displays a surprisingly accessible
explanation of the inconsistency, which is philosophically profound
and never presented in the literature. The detection of this
inconsistency in combination with the work reported here thus
demonstrates that artificial intelligence systems -- particularly
higher-order automated theorem provers -- are capable of assisting in
the discovery and elucidation of \emph{new} and philosophically
relevant knowledge.

@@

\subsection{Related Work}

First successful applications of theorem proving technology in
metaphysics were reported by Fitelson, Oppenheimer and
Zalta~\cite{FitelsonZalta,oppenheimer11}, who coined the term \textit{Computational Metaphysics} for this new research area and employed the first-order
\textsc{Prover9} \cite{prover9-mace4} in their experiments. Later on,
Rushby~\cite{rushby13} used the proof assistant \textsc{PVS} \cite{cade92-pvs}. Common to both
works is a significant amount of proof-hand-coding work as well as their
focus on a non-modal formalization of St. Anselm's~\cite{Proslogion} simpler 
and older ontological argument. In contrast, the greater complexity of G\"odel's argument requires the formalization and automation of variants of \emph{higher-order} and \emph{modal} logics.

\begin{wrapfigure}[22]{r}{0.55\textwidth} \centering \vskip-3em
  \includegraphics[width=.52\textwidth]{./Scott_S5U.png}
  \caption{Full Automation of T3 in \SFiveU; Consistency of Scott's
  Axioms;  Automatic Verification of Modal Collapse} \label{Scott_S5U}
\end{wrapfigure} 
\section{A Brief History}\label{sec:history}
St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz~\cite{Adams}.  
Although G\"odel's notion of positive property is not exactly the same
as Leibniz's notion of perfection, G\"odel's manuscript
(Fig.~\ref{GoedelScript}) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5U}. 
The main difference to G\"odel's version is an
extra conjunct in the definition of \emph{essence} (\emph{ess}). G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

\begin{wrapfigure}[20]{r}{0.55\textwidth}
    \centering \vskip-1em
    \centerline{\includegraphics[width=0.52\textwidth]{./InconsistencyIsabelleK.png}}
 \caption{Inconsistency of G\"odel's Axioms in HOML \K verified in Isabelle/HOL} \label{InconsistencyIsabelleK}
\end{wrapfigure}
G\"odel's omission has been
considered inessential and merely an oversight by many. For instance,
Hazen [\cite[p.365]{Hazen}] states that ``G\"odel left this
clause out [\ldots] but this appears to have been an oversight''.
% -- it is
%included in related manuscripts''. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analysis and criticisms of the
argument. Especially since the discovery by Sobel
\cite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well.\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; they must be analyzed
  separately.} Opponents of the argument
(e.g. Oppy [\cite[p.226--227]{oppy96:_goedel_ontol_argum};
\cite[p.364]{oppy00:_respon_gettin};
\cite[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

Logic textbooks % \cite{ToDo:which}
commonly utilize higher-order logic in an informal/semi-formal way as
a meta-language to introduce the syntax and the semantics of object
logics of interest, in which reasoning problems in concrete
application domains can be modeled and solved with pen and paper. In
fact, this approach can also be followed on the computer (using HOL as
a formal meta-language) for even very challenging object logics (such
as HOML) to enable interactive and automated theorem proving with
existing theorem provers for HOL.

@@

For a computational analysis of G\"odel's ontological argument, the embedding of HOMLs such as \textbf{K},
\textbf{KB} and \textbf{S5} with various domain conditions (possibilist and actualist quantification)
is required. This idea has been successfully employed in related work
\cite{C40}. The embedding of HOML is in fact
straightforward. Formulas in HOML are \emph{lifted}, i.e., converted into predicates
over worlds, which are themselves explicitly represented as
terms. The logical constants of HOML are translated to HOL terms in such a way that, for instance, 
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation
\cite{DBLP:journals/logcom/Ohlbach91}, which is here intra-logically
realized --- and extended for quantifiers --- in HOL by stating a set
of equations defining the logical constants (Fig.~\ref{QML_S5U}). The
resulting object logic is the HOML \textbf{K} with rigid terms and constant
domains (possibilist quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) are
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by using
 an existence predicate to guard the
quantifiers. The embedding approach is, therefore, very flexible.

@@

In the typical workflow during an attempt to prove a conjecture with a theorem prover, it is customary to check the consistency of the axioms first. For if the axioms are inconsistent, anything (including the conjecture) would be trivially derivable in classical logic (\emph{ex falso quodlibet}). Surprisingly, when this routine check was performed on G\"odel's axioms \cite{C40}, the \textsc{Leo-II} prover claimed that the axioms were inconsistent. Unfortunately, the refutation generated by \textsc{Leo-II} was barely human-readable. The text file was 153 lines\footnote{Long lines with an average of 184 characters per line.} long and used machine-oriented calculus (higher-order resolution \cite{W47}) and syntax (TPTP THF \cite{J22}). Part of the file is displayed in Fig.~\ref{LEO-Proof}.

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{./LEO-Proof.png}}
\caption{Lines 115--120 of \textsc{Leo-II}'s refutation. Primitive
  substitutions (e.g. with the empty property) are highlighted. In the
red part (see $\boldsymbol{\leftarrow}$), property variable $\texttt{SV8}$ has been instantiated with the
$\lambda \texttt{SV16}_{\mu}. \lambda \texttt{SV17}_{\iota}. \bot$, i.e., the (lifted) empty property.
} \label{LEO-Proof}
\end{figure*}

@@

\subsection{Informal Argument} \label{sec:arg1}
Based on the hints found in \textsc{Leo-II}'s refutation, we conceived the following informal explanation for the inconsistency of G\"odel's axioms:

\begin{enumerate}
\item From G\"odel's definition of essence 
(${\ess{\phi}{x} \biimp {\allq \psi} (\psi(x)
\imp {\nec} \allq y (\phi(y) \imp \psi(y)))}$) it follows that the
empty property (or self-difference) is an essence of every individual (\textbf{Empty Essence Lemma}): 
$$\allq x\; (\ess{\emptyset}{x})$$

\item From theorem T1 (\textit{Positive properties are possibly
  exemplified}: ${\allq \phi} [P(\phi) \imp {\pos}  \exq x
  \phi(x)]$) and axiom A5 (``necessary existence'' is a positive property: $P(\NE)$ ), it follows that $\NE$ is possibly exemplified:
  $$
  \pos \exq x [\NE(x)]
  $$

\item Expanding the definition of ``necessary existence''
  (${\NE(x) \equiv \allq \phi [\ess{\phi}{x} \imp \nec \exq y
    \phi(y)]}$), the following is obtained:
  $$
  \pos \exq x [\allq \varphi [ \ess{\varphi}{x} \imp \nec \exq y [\varphi(y)] ] ]
  $$

\item The sentence above holds for all $\varphi$ and thus, in
  particular, for the empty property (or self-difference):
$$
\pos \exq x [ \ess{\emptyset}{x} \imp \nec \exq y [\emptyset(y)] ]
$$

\item By the Empty Essence Lemma, the antecedent of the implication above is valid. Therefore, the sentence above entails:
$$
\pos \exq x [ \nec \exq y [\emptyset(y)] ]
$$

\item By definition of $\emptyset$: 
$$
\pos \exq x [ \nec \bot ]
$$

\item As the existential quantifier is binding no variable within its scope, the sentence is equi-valid with:
$$\pos \nec \bot $$

@@

\subsection{Argument Reconstruction in Isabelle}  \label{sec:arg2}
\begin{figure}[t]
\centerline{\includegraphics[width=1\columnwidth]{./InconsistencyIsabelleK.png}}
\caption{Inconsistency of G\"odel's Axioms in HOML \K verified in Isabelle/HOL} \label{InconsistencyIsabelleK}
\end{figure}
\begin{figure*}[t]
  \centering
  \begin{subfigure}[t]{0.715\textwidth}
    \includegraphics[width=\textwidth]{./Manuscript2.png}
    \caption{G\"{o}del's manuscript, with mutually inconsistent axioms and
      definitions highlighted (with permission from the Kurt G\"odel Papers, Shelby White and Leon Levy Archives Center, Princeton, NJ, USA, on deposit at Princeton University)} \label{GoedelScript} 
  \end{subfigure}
   \begin{subfigure}[t]{0.28\textwidth}
     \includegraphics[width=\textwidth,height=7.2cm]{./Inconsistency_S5U_direct.png}
     \caption{Inconsistency in HOML \SFiveU} \label{Inconsistency_S5U} 
   \end{subfigure}
% \includegraphics[width=.73\textwidth]{./Images/Manuscript2.png} \hfill
% \includegraphics[width=.26\textwidth]{./Images/Inconsistency_S5U_direct.png}
 \caption{The inconsistency in G\"{o}del's manuscript has been
   detected and verified by HOL ATPs} 
\end{figure*}
To verify the correctness of the informal argument explained above, it
was reconstructed in Isabelle/HOL, using Metis\footnote{Metis, unlike
  external provers such as \textsc{Leo-II} or Satallax, 
  constructs proofs in Isabelle's highly trusted kernel calculus.} to automate the
inessential parts (cf. Fig.~\ref {InconsistencyIsabelleK}). The essential use of the Empty Essence Lemma, on
the other hand, is explicitly stated, to ensure that Isabelle is
reconstructing the same argument. In fact, without the help of this
lemma, Metis is still not strong enough to refute G\"odel's
axioms.

\subsection{Mapping the Inconsistency to G\"odel}

The inconsistency verified in Fig.~\ref{InconsistencyIsabelleK} follows from the definition of
essence (ess), the definition of necessary existence (NE), the
axioms A1a and A2 (which entail theorem T1), and axiom A5. It remains to show that
these ingredients are actually present in G\"odel's manuscript in
Fig.~\ref{GoedelScript}.

@@

\subsection{Inconsistency of G\"odel's Axioms in \SFiveU}

Isabelle/HOL's Sledgehammer tool, which orchestrates calls to
external provers such as \textsc{Leo-II}, still
fails to detect the inconsistency of G\"odel's axioms in the standard
embedding of \SFive, while a direct modeling of the problem in TPTP THF syntax
in combination with a direct call of \textsc{Leo-II} succeeded. In
other words, without independent experiments with no mediation through Sledgehammer, the
inconsistency would not have been detected.

@@

\pagebreak
%A discussion between Chad Brown and Benzm\"uller
% significantly influenced the rational reconstruction of the inconsistency reported by \textsc{LEO-II}.

>>>>

\title{The Inconsistency in G\"odel's Ontological Argument: A Success
  Story for Automated Theorem Proving in Metaphysics}
\author{Christoph Benzm\"uller\inst{1}\thanks{This work has been supported by
    the German Research Foundation DFG under grants BE2501/9-1,2} \and Bruno Woltzenlogel Paleo\inst{2} 
}

@@

\begin{abstract}
  We discusses the discovery of the inconsistency in G\"odel's
  ontological argument as a success story for artificial
  intelligence. Despite the popularity of the argument since the
  appearance of G\"odel's manuscript in the early 1970's, the
  inconsistency of the axioms used in the argument remained unnoticed
  until 2013, when it was detected automatically by the higher-order
  theorem prover \textsc{Leo-II}.  Understanding and verifying the
  refutation generated by the prover turned out to be a time-consuming
  task. Its completion, as reported here, required the
  reconstruction of the refutation in the Isabelle proof assistant.% , and it
  % also led to a novel and more efficient way of automating
  % higher-order modal logic \textbf{S5} with a universal accessibility
  % relation. Furthermore, the development of an improved syntactical
  % hiding for the utilized logic embedding technique allows the refutation to be
  % presented in a human-friendly way, suitable for non-experts in the
  % technicalities of higher-order theorem proving. This brings us a
  % step closer to wider adoption of logic-based artificial intelligence
  % tools by philosophers.

  (This paper is a short version of our contribution to IJCAI 2016.)
\end{abstract}

@@

The main novel contribution is a detailed
analysis (in various modal logics) of the inconsistency of G\"{o}del's
original version of the axioms used in his manuscript
\cite{GoedelNotes}. The extraction, reconstruction and
verification of an informal, human intuitive argument has been an open
problem since the first detection of this inconsistency by
the
\textsc{Leo-II} prover \cite{C40}.  The verified refutation (discussed in
\S\ref{sec:inconsistency}) displays a surprisingly accessible
explanation of the inconsistency, which is philosophically profound
and never presented in the literature. The detection of this
inconsistency in combination with the work reported here thus
demonstrates that 
higher-order automated theorem provers are capable of assisting in
the discovery and elucidation of \emph{new} and philosophically
relevant knowledge.

@@

\paragraph{Related Work.}
First successful applications of theorem proving technology in
metaphysics were reported by Fitelson, Oppenheimer and
Zalta~\cite{FitelsonZalta,oppenheimer11}, who coined the term \textit{Computational Metaphysics} for this new research area and employed the first-order
\textsc{Prover9} \cite{prover9-mace4} in their experiments. Later on,
Rushby~\cite{rushby13} used the proof assistant \textsc{PVS} \cite{cade92-pvs}. Common to both
works is a significant amount of proof-hand-coding work as well as their
focus on a non-modal formalization of St. Anselm's~\cite{Proslogion} simpler 
and older ontological argument. In contrast, the greater complexity of G\"odel's argument requires the formalization and automation of variants of \emph{higher-order} and \emph{modal} logics.

\begin{wrapfigure}[22]{r}{0.55\textwidth} \centering \vskip-3em
  \includegraphics[width=.52\textwidth]{./Scott_S5U.png}
  \caption{Full Automation of T3 in \SFiveU; Consistency of Scott's
  Axioms;  Automatic Verification of Modal Collapse} \label{Scott_S5U}
\end{wrapfigure} 
\section{A Brief History}\label{sec:history}
St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz~\cite{Adams}.  
Although G\"odel's notion of positive property is not exactly the same
as Leibniz's notion of perfection, G\"odel's manuscript
(Fig.~\ref{GoedelScript}) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5U}. 
The main difference to G\"odel's version is an
extra conjunct in the definition of \emph{essence} (\emph{ess}). G\"odel's
different definition of essence can be seen 
% either in his manuscript
% (Fig.~\ref{GoedelScript}) or, in more modern notation,
in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

\begin{wrapfigure}[20]{r}{0.55\textwidth}
    \centering \vskip-1em
    \centerline{\includegraphics[width=0.52\textwidth]{./InconsistencyIsabelleK.png}}
 \caption{Inconsistency of G\"odel's Axioms in HOML \K verified in Isabelle/HOL} \label{InconsistencyIsabelleK}
\end{wrapfigure}
G\"odel's omission has been
considered inessential and merely an oversight by many. 
% For instance,
% Hazen [\cite[p.365]{Hazen}] states that ``G\"odel left this
% clause out [\ldots] but this appears to have been an oversight''.
% -- it is
%included in related manuscripts''. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analysis and criticisms of the
argument. Especially since the discovery by Sobel
\cite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well.\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; they must be analyzed
  separately.} Opponents of the argument
(e.g. Oppy [\cite[p.226--227]{oppy96:_goedel_ontol_argum};
\cite[p.364]{oppy00:_respon_gettin};
\cite[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

% Logic textbooks % \cite{ToDo:which}
% commonly utilize higher-order logic in an informal/semi-formal way as
% a meta-language to introduce the syntax and the semantics of object
% logics of interest, in which reasoning problems in concrete
% application domains can be modeled and solved with pen and paper. In
% fact, this approach can also be followed on the computer (using HOL as
% a formal meta-language) for even very challenging object logics (such
% as HOML) to enable interactive and automated theorem proving with
% existing theorem provers for HOL.

@@

For a computational analysis of G\"odel's ontological argument, we
utilised an embedding of HOMLs, such as \textbf{K}, \textbf{KB} and
\textbf{S5} with various domain conditions (possibilist and actualist
quantification), in HOL. This idea has already been successfully
employed in earlier work \cite{C40}. The embedding of HOML is in fact
straightforward. Formulas in HOML are \emph{lifted}, i.e., converted
into predicates over worlds, which are themselves explicitly
represented as terms. The logical constants of HOML are translated to
HOL terms in such a way that, for instance,
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation,
%\cite{DBLP:journals/logcom/Ohlbach91}, 
which is here intra-logically realized --- and extended for
quantifiers --- in HOL by stating a set of equations defining the
logical constants. The resulting object logic is
the HOML \textbf{K} with rigid terms and constant domains (possibilist
quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) are
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
using an existence predicate to guard the quantifiers. The embedding
approach is, therefore, very flexible.

@@

In the typical workflow during an attempt to prove a conjecture with a
theorem prover, it is customary to check the consistency of the axioms
first. For if the axioms are inconsistent, anything (including the
conjecture) would be trivially derivable in classical logic (\emph{ex
  falso quodlibet}). Surprisingly, when this routine check was
performed on G\"odel's axioms \cite{C40}, the \textsc{Leo-II} prover
claimed that the axioms were inconsistent. Unfortunately, the
refutation generated by \textsc{Leo-II} was barely human-readable. The
text file was 153 lines\footnote{Long lines with an average of 184
  characters per line.} long and used machine-oriented calculus
(higher-order resolution \cite{W47}) and syntax (TPTP THF
\cite{J22}).

% Part of the file is displayed in Fig.~\ref{LEO-Proof}.

% \begin{figure*}
% \centerline{\includegraphics[width=\textwidth]{./LEO-Proof.png}}
% \caption{Lines 115--120 of \textsc{Leo-II}'s refutation. Primitive
%   substitutions (e.g. with the empty property) are highlighted. In the
% red part (see $\boldsymbol{\leftarrow}$), property variable $\texttt{SV8}$ has been instantiated with the
% $\lambda \texttt{SV16}_{\mu}. \lambda \texttt{SV17}_{\iota}. \bot$, i.e., the (lifted) empty property.
% } \label{LEO-Proof}
% \end{figure*}

@@

\paragraph{Informal Argument.} \label{sec:arg1}
Based on the hints found in \textsc{Leo-II}'s refutation, we conceived the following informal explanation for the inconsistency of G\"odel's axioms:

\begin{enumerate}
\item From G\"odel's definition of essence 
(${\ess{\phi}{x} \biimp {\allq \psi} (\psi(x)
\imp {\nec} \allq y (\phi(y) \imp \psi(y)))}$) it follows that the
empty property (or self-difference) is an essence of every individual
(\textbf{Empty Essence Lemma}): \hfill $\allq x\; (\ess{\emptyset}{x})$

\item From theorem T1 (\textit{Positive properties are possibly
  exemplified}: ${\allq \phi} [P(\phi) \imp {\pos}  \exq x
  \phi(x)]$) and axiom A5 (``necessary existence'' is a positive
  property: $P(\NE)$ ), it follows that $\NE$ is possibly exemplified:
  \hfill $  \pos \exq x [\NE(x)] $

\item Expanding the definition of ``necessary existence''
  (${\NE(x) \equiv \allq \phi [\ess{\phi}{x} \imp \nec \exq y
    \phi(y)]}$), the following is obtained: \hfill $  \pos \exq x
  [\allq \varphi [ \ess{\varphi}{x} \imp \nec \exq y [\varphi(y)] ] ] $

\item The sentence above holds for all $\varphi$ and thus, in
  particular, for the empty property (or self-difference): \hfill $\pos \exq x [ \ess{\emptyset}{x} \imp \nec \exq y [\emptyset(y)] ]$

\item By the Empty Essence Lemma, the antecedent of the implication
  above is valid. Therefore, the sentence above entails: \hfill $\pos \exq x [ \nec \exq y [\emptyset(y)] ]$

\item By definition of $\emptyset$: \hfill $\pos \exq x [ \nec \bot ]$

\item As the existential quantifier is binding no variable within its
  scope, the sentence is equi-valid with: \hfill $\pos \nec \bot $

@@

\paragraph{Argument Reconstruction in Isabelle.}  \label{sec:arg2}
\begin{figure*}[t]
  \centering
  \begin{subfigure}[t]{0.715\textwidth}
    \includegraphics[width=\textwidth]{./Manuscript2.png}
    \caption{G\"{o}del's manuscript, with mutually inconsistent axioms and
      definitions highlighted (with permission from the Kurt G\"odel Papers, Shelby White and Leon Levy Archives Center, Princeton, NJ, USA, on deposit at Princeton University)} \label{GoedelScript} 
  \end{subfigure}
   \begin{subfigure}[t]{0.28\textwidth}
     \includegraphics[width=\textwidth,height=7.2cm]{./Inconsistency_S5U_direct.png}
     \caption{Inconsistency in HOML \SFiveU} \label{Inconsistency_S5U} 
   \end{subfigure}
% \includegraphics[width=.73\textwidth]{./Images/Manuscript2.png} \hfill
% \includegraphics[width=.26\textwidth]{./Images/Inconsistency_S5U_direct.png}
 \caption{The inconsistency in G\"{o}del's manuscript has been
   detected and verified by HOL ATPs} 
\end{figure*}
To verify the correctness of the informal argument explained above, it
was reconstructed in Isabelle/HOL, using Metis\footnote{Metis, unlike
  external provers such as \textsc{Leo-II} or Satallax, 
  constructs proofs in Isabelle's highly trusted kernel calculus.} to automate the
inessential parts (cf. Fig.~\ref {InconsistencyIsabelleK}). The essential use of the Empty Essence Lemma, on
the other hand, is explicitly stated, to ensure that Isabelle is
reconstructing the same argument. In fact, without the help of this
lemma, Metis is still not strong enough to refute G\"odel's
axioms.

\paragraph{Mapping the Inconsistency to G\"odel.}
The inconsistency verified in Fig.~\ref{InconsistencyIsabelleK} follows from the definition of
essence (ess), the definition of necessary existence (NE), the
axioms A1a and A2 (which entail theorem T1), and axiom A5. It remains to show that
these ingredients are actually present in G\"odel's manuscript in
Fig.~\ref{GoedelScript}.

@@

\paragraph{Inconsistency of G\"odel's Axioms in \SFiveU.}
Isabelle/HOL's Sledgehammer tool, which orchestrates calls to
external provers such as \textsc{Leo-II}, still
fails to detect the inconsistency of G\"odel's axioms in the standard
embedding of \SFive, while a direct modeling of the problem in TPTP THF syntax
in combination with a direct call of \textsc{Leo-II} succeeded. In
other words, without independent experiments with no mediation through Sledgehammer, the
inconsistency would not have been detected.

@@

%\pagebreak
%A discussion between Chad Brown and Benzm\"uller
% significantly influenced the rational reconstruction of the inconsistency reported by \textsc{LEO-II}.
